Missing data in sports science: Current practice in football research and recommendations for reporting missingness

### 2020-08-28 paper/paper.Rmd
updates references and changes "problem" to "case study"

Missing data can undermine the validity of research results
[@Sainani2015; @Nakagawa2008; Sterne2009]. Data can go
missing many ways. For example, accidentally skipping a survey question,
equipment failure, or intentionally not recording values. Before
analysis, missing data must be handled, with values typically deleted or
imputed. There is no universal approach to handling missing data with
contextual factors -- such as, the study design and objective, and
pattern of missingness -- determining how to handle missings on a
case-by-case basis [@Sainani2015].
The effect of missing data on the validity and accuracy of a study's
results is largely determined by the reasons for missingness, and how
missing data are handled.

@@

Problem: *Unreported missing values*. While not always the case, even if
not reported, missing data can be obvious. A recent paper examined the
relationship between match performance indicators and outcome, in
Australian Football between the 2001 and 2016. Problematically, at least
one of the 91 team performance indicators included in the analysis was
not available over the entire 2001-2016 period [@Young2019]. The variable
'meters gained' was not collected over the specified duration [@ref].
This was not noted anywhere in the paper, or in the supplementary
materials. It is unknown whether imputation was undertaken. Without
these details, it is difficult for other analysts and researchers to use
and/or extend the ideas presented in the paper, or reproduce the
analysis. It is also difficult for a reader to evaluate how the
missingness could bias the results. Unfortunately, the absence of a
missing data statement when data are missing is not unique to the study
above [@Young2019]. Authors should include a missing data statement,
irrespective of whether data is missing, or not.

Problem: *Mean imputation.* The default of most statistical software for
handling missing data in modelling is listwise deletion. Listwise
deletion removes entire rows that contain missing observations from the
analysis. At best, listwise deletion reduces the power in analysis by
reducing the sample size. At worst, it introduces bias. For example, if
injured participants are removed from a study on injury prevention.
Imputing values removes the need to conduct listwise deletion. However,
it is critical how these values are imputed, and that the imputation
method is documented. One imputation method is imputing the mean value
(e.g., participant or group average) [@Young2019]. While imputing the mean
preserves the study sample size (and point estimates), it also reduces
the variance, can alter the relationship between the variable with
missing observations and other variables, and can bias (underestimate)
standard errors [@Scheffer2002].
Smaller standard errors typically reduce *p*-values, which may lead to
incorrect inference. Mean imputation should be avoided. An imputation
strategy that incorporates information from other related variables in
the data -- such as linear regression, or K nearest neighbours -- should
be considered. This is discussed in the recommendations section.

Problem: *Not evaluating the effect, or choice, of imputation.* The
imputation method used to impute missing data has the potential to
affect the outcome of a study. Therefore, it is important that authors
understand, and document, how imputation affected the goal of the
analysis (i.e., inference, prediction, or both). For example,
understanding the implications in using mean imputation, compared to
K-nearest neighbour imputation -- the effect on parameter estimates
('significant' versus 'not significant'), and the (un)certainty of the
coefficients (i.e., less/more). When using imputation, authors need to
perform a sensitivity analysis (ref circulation article-Nick). This is
discussed below.

@@

# Supplement 1

# Supplement 2

# Supplement 3

>>>>

Missing data can undermine the validity of research results
[@Sainani2015; @Nakagawa2008; @Sterne2009]. Data can go
missing many ways. For example, accidentally skipping a survey question,
equipment failure, or intentionally not recording values. Before
analysis, missing data must be handled, with values typically deleted or
imputed. There is no universal approach to handling missing data with
contextual factors -- such as, the study design and objective, and
pattern of missingness -- determining how to handle missings on a
case-by-case basis [@Sainani2015].
The effect of missing data on the validity and accuracy of a study's
results is largely determined by the reasons for missingness, and how
missing data are handled.

@@

Case study: *Unreported missing values*. While not always the case, even if
not reported, missing data can be obvious. A recent paper examined the
relationship between match performance indicators and outcome, in
Australian Football between the 2001 and 2016. Problematically, at least
one of the 91 team performance indicators included in the analysis was
not available over the entire 2001-2016 period [@Young2019]. The variable
'meters gained' was not collected over the specified duration [@ref].
This was not noted anywhere in the paper, or in the supplementary
materials. It is unknown whether imputation was undertaken. Without
these details, it is difficult for other analysts and researchers to use
and/or extend the ideas presented in the paper, or reproduce the
analysis. It is also difficult for a reader to evaluate how the
missingness could bias the results. Unfortunately, the absence of a
missing data statement when data are missing is not unique to the study
above [@Young2019]. Authors should include a missing data statement,
irrespective of whether data is missing, or not.

Case study: *Mean imputation.* The default of most statistical software for
handling missing data in modelling is listwise deletion. Listwise
deletion removes entire rows that contain missing observations from the
analysis. At best, listwise deletion reduces the power in analysis by
reducing the sample size. At worst, it introduces bias. For example, if
injured participants are removed from a study on injury prevention.
Imputing values removes the need to conduct listwise deletion. However,
it is critical how these values are imputed, and that the imputation
method is documented. One imputation method is imputing the mean value
(e.g., participant or group average) [@Young2019]. While imputing the mean
preserves the study sample size (and point estimates), it also reduces
the variance, can alter the relationship between the variable with
missing observations and other variables, and can bias (underestimate)
standard errors [@Scheffer2002].
Smaller standard errors typically reduce *p*-values, which may lead to
incorrect inference. Mean imputation should be avoided. An imputation
strategy that incorporates information from other related variables in
the data -- such as linear regression, or K nearest neighbours -- should
be considered. This is discussed in the recommendations section.

Case study: *Not evaluating the effect, or choice, of imputation.* The
imputation method used to impute missing data has the potential to
affect the outcome of a study. Therefore, it is important that authors
understand, and document, how imputation affected the goal of the
analysis (i.e., inference, prediction, or both). For example,
understanding the implications in using mean imputation, compared to
K-nearest neighbour imputation -- the effect on parameter estimates
('significant' versus 'not significant'), and the (un)certainty of the
coefficients (i.e., less/more). When using imputation, authors need to
perform a sensitivity analysis (ref circulation article-Nick). This is
discussed below.

@@
